
\documentclass{report}
    \title{COMP70015 Mathematics for Machine Learning notes}
    \author{Timothy Chung}
    \date{16/05/24}
    
    \usepackage[a4paper, total={7in, 10in}]{geometry}
    \setlength\parindent{0pt}
    \input{common.tex}
    \newcommand{\infer}[2]{\cfrac{#1}{#2}}
    \newcommand{\infers}[3]{\cfrac{#1}{#2}(#3)}
    
    \newcommand{\ninfer}[3]{(\text{#1}):\cfrac{#2}{#3}}
    \newcommand{\ninfers}[4]{(\text{#1}):\cfrac{#2}{#3}(#4)}
    \newcommand{\ainfer}[3]{\cfrac{#2}{#3}(\text{#1})}
    
    \newcommand{\mllet}[2]{\text{let } #1 \text{ in } #2}
    \newcommand{\mlfix}[2]{\text{fix } #1 . #2}
    
    \newcommand{\hc}{\mathcal{H}}
    \newcommand{\cov}{\text{Cov}}
    \newcommand{\E}{\text{E}}
    \newcommand{\var}{\text{Var}}
    \newcommand{\mb}{\mathbf}
    \newcommand{\est}{\hat{\theta}}
    \newcommand{\p}{\theta}
    
    
    
\begin{document}
\include{titlepage/titlepage}
\setlength{\cftsecnumwidth}{4.2em} 

\begin{titlepage}
\centering
    \vspace*{1cm}
    \Huge
    \textbf{COMP70015 Mathematics for Machine Learning} \\
    \vspace{1cm}
    \Large
    Timothy Chung \\
    \vspace{1cm}
    Autumn 2025 \\
    \vfill
\end{titlepage}

\setcounter{tocdepth}{1}
\tableofcontents
\newpage
\setcounter{chapter}{-1}
\chapter{Foreword}
The notes are an amalgamation of the course's informal notes with some additions from the slides and any source material referenced.
\section{Box Previews}

\begin{definitionbox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.
\end{definitionbox}

\begin{sidenotebox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.
\end{sidenotebox}

\begin{examplebox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.
\end{examplebox}

\begin{exambox}{1c}{2018}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.
\end{exambox}


\begin{commentbox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.
\end{commentbox}

\begin{theorembox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.    
\end{theorembox}

\begin{referencebox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.    
\end{referencebox}

\begin{intuitbox}{Machine Learning}
    Machine learning is the study of computer algorithms that improve automatically through experience. It is seen as a subset of artificial intelligence. Machine learning algorithms build a mathematical model based on sample data, known as "training data", in order to make predictions or decisions without being explicitly programmed to perform the task.    
\end{intuitbox}

\input{1_intro}


\end{document}

